Conditional Power and Friends: The Why and How of (Un)planned, Unblinded Sample Size Recalculations in Confirmatory Trials

### 2020-10-13 latex/manuscript.tex
Updates from Overleaf

\author{
  Kevin Kunzmann \thanks{
        MRC Biostatistics Unit, University of Cambridge,
        Cambridge Institute of Public Health,
        Forvie Site, Robinson Way,
        Cambridge Biomedical Campus,
        Cambridge CB2 0SR,
        United Kingdom
    }\\
    MRC Biostatistics Unit \\
    University of Cambridge \\
    \texttt{kevin.kunzmann@mrc-bsu.cam.ac.uk} \\

    Michael J. Grayling \\
    Population Health Sciences Institute \\
    Newcastle University \\
    \texttt{michael.grayling@newcastle.ac.uk} \\

    Kim May Lee \\
    Pragmatic Clinical Trials Unit\\
    Queen Mary University of London\\
    \texttt{k.m.lee@qmul.ac.uk} \\

    David S.\ Robertson \\
    MRC Biostatisics Unit\\
    University of Cambridge\\
    \texttt{david.robertson@mrc-bsu.cam.ac.uk} \\

    Kaspar Rufibach \\
    Methods, Collaboration, and Outreach Group (MCO) \\
    Department of Biostatistics\\
    F. Hoffmann-La Roche, Basel\\
    \texttt{kaspar.rufibach@roche.com} \\

    James M. S. Wason \\
    Population Health Sciences Institute \\
    Newcastle University\\
    and \\
     MRC Biostatistics Unit \\
    University of Cambridge \\
    \texttt{james.wason@newcastle.ac.uk}
}

@@

The three estimators are depicted in Figure~\ref{fig:acp-ocp-pp-comparison-absolute} as functions 
of the observed interim outcome, where the interim time-point $m = 26$.
\begin{figure}
    \centering
    \begin{subfigure}[b]{\textwidth}
        \centering
        \includegraphics[width=.9\textwidth]{figures/acp_ocp_pp_example}
        \caption{%
            Estimates as function of the interim data.
        }
        \label{fig:acp-ocp-pp-comparison-absolute}
    \end{subfigure}
    \begin{subfigure}[b]{\textwidth}
        \centering
        \includegraphics[width=.9\textwidth]{latex/figures/acp_ocp_pp_sampling_distributions}
        \caption{%
            Histograms of the sampling distributions.
        }
        \label{fig:sampling-distributions}
    \end{subfigure}
    \begin{subfigure}[b]{\textwidth}
        \centering
        \includegraphics[width=.9\textwidth]{latex/figures/acp_ocp_pp_bias_mae_mse}
        \caption{%
            Bias, mean absolute error (MAE) and mean squared error (MSE).
        }
        \label{fig:acp-ocp-pp-as-estimators}
    \end{subfigure}
    \caption{%
        Properties of $\ACP,\OCP$, and $\PP$ as estimators of the unknown conditional power at $m=26$ an overall sample size $n=79$.
    }
\end{figure}
The most sensitive measure is $\OCP$ while $\ACP$ is the least sensitive
(see also spread of the sampling distributions in Figure~\ref{fig:sampling-distributions}). 
This is due to $\ACP$ assuming a fixed effect size and thus 
only being affected by direct changes in $z_m$.
Both $\OCP$ and $\PP$, however, are also indirectly affected by updating 
the belief about the effect size with the interim results.
The plug-in estimate $\OCP$ assumes that the observed effect is the 
true effect and does not quantify the uncertainty around this value in any way. 
$\PP$, on the other hand, invokes Bayes' theorem and then integrates 
$\CP(z_m,c,\theta)$ with respect to the obtained posterior distribution. 
The degree of adaptation of $\PP$ thus depends 
on the vagueness of the chosen prior. 
In fact, as the prior approaches a point mass at $\theta_1$, $\PP$ converges
to $\ACP$.
$\ACP$ can thus be seen as a mere special case of $\PP$ with a point prior
on $\theta_1>0$.

@@

If the recalculation rule defined in \eqref{eq:naive-recalc}--\eqref{eq:nmax-naive-recalc} is made mandatory in the
study protocol, 
it yields two functions $n(z_m)$ and $c(z_m)$ as the point-wise solution of the problem
which jointly define an `adaptive design' for any $z_m$.
Here, `mandatory' means that it is decided \textit{a~priori} to always (for all $z_m$)
recalculate the 
final sample size and critical value in the 
above specified way after a fixed number of outcomes $m$ has been observed.
The term `adaptive design' is slightly misleading in this context.
The design itself is pre-specified (and thus not adapted or changed) 
but only $n$ and $c$ are adaptive since they vary as functions of $z_m$.
We would like to emphasize that such a detailed specification of the sample size recalculation rule is not common in practice since most protocols are focused on guaranteeing control of the maximal type-I-error rate control and maintaining maximal flexibility during the interim analysis.
Still, if a sample-size reassessment was prescribed in the study protocol, there must have been some rationale as to the goal of that reassessment and we merely investigate the consequences for a design's unconditional properties when applying such a rationale in a mandatory way. 
The described approach should thus be seen as a hypothetical example highlighting the ineffectiveness that can arise from a na\"{\i}ve implementation of a recalculation based on a fixed threshold for 
conditional power.
The corresponding sample size and critical value functions for $m=26$ are depicted in Figure~\ref{fig:binding-recalculation}.
\begin{figure}
    \centering
    \includegraphics[width=\textwidth]{latex/figures/naive_adaptive_and_optimal_designs}
    \caption{%
        Original single-stage design; na\"{\i}vely adapted design 
        with sample size $n'(z_m)$ and critical value $c'(z_m)$ functions defined by the conditional optimization problem \eqref{eq:naive-recalc}--\eqref{eq:nmax-naive-recalc} and 
        ${n=79}, {m=26}, {n_{max}=160}, {n_{min}=30}$ and $1-\beta_{cond.}=0.8$;
        the optimal design is the solution of \eqref{eq:optimal-design-start}--\eqref{eq:optimal-design-end}.
    }
    \label{fig:binding-recalculation}
\end{figure}

@@

We illustrate both approaches by example of the optimal two-stage design that minimizes the objective 
$\E_\varphi[\,n(Z_m)\,]$ subject to a maximal type-I error rate of $0.025$ and a minimal expected power of $0.8$, see~Figure~\ref{fig:binding-recalculation}.  
This design has a pre-planned interim analysis after $m=35$ individuals.
Now assume that
the prior $\varphi$ needs to be revised due to new trial-external 
information at the pre-planned interim analysis.
We consider members of the class of shifted truncated Normal distributions 
\begin{align}
    \psi_{\mu}(\theta) := \frac{\displaystyle \boldsymbol{1}_{[-0.5, 1]}(\theta)\,\phi\left(\frac{\theta - \mu}{0.2}\right)}{\displaystyle \Phi\left(\frac{1 - \mu}{0.2}\right) - \Phi\left(\frac{-0.5 - \mu}{0.2}\right)}
\end{align}
as revised priors, i.e. the original prior~$\varphi=\psi_{0.4}$.
Figure~\ref{fig:recalculation-old-m} shows the behavior of the recalculated sample size and critical value in this scenario for observed effects of $z_m/\sqrt{m}=0.25, 0.3, 0.35$ and varying mean of the revised prior if the originally planned interim time-point is maintained.
\begin{figure}
    \centering
    \begin{subfigure}[b]{\textwidth}
        \centering
        \includegraphics[width=.9\textwidth]{latex/figures/recalculation-type-II-original-m}
        \caption{%
            Interim analysis at the originally planned interim sample size of $m=35$ for varying observed effect.
        }
        \label{fig:recalculation-old-m}
    \end{subfigure}
    \begin{subfigure}[b]{\textwidth}
        \centering
        \includegraphics[width=.9\textwidth]{latex/figures/recalculation-type-II-earlier}
        \caption{%
            Earlier-than-planned interim analysis at an observed effect of $0.3$ and varying interim time-point.
        }
        \label{fig:recalculation-earlier}
    \end{subfigure}
    \begin{subfigure}[b]{\textwidth}
        \centering
        \includegraphics[width=.9\textwidth]{latex/figures/recalculation-lambda-approach-original-m}
        \caption{%
            Interim analysis at the originally planned interim sample size of $m=35$ for varying observed effect and the $\lambda$ approach.
        }
        \label{fig:recalculation-lambda}
    \end{subfigure}
    \caption{%
        Comparison of consistent unplanned recalculation schemes.
    }
\end{figure}
Clearly, the original design is indeed invariant if the prior remains unchanged.
The sample size is increased if the new prior is more conservative
than $\varphi$ and \textit{vice~versa}.

>>>>

\author{
  Kevin Kunzmann \thanks{
        MRC Biostatistics Unit, University of Cambridge,
        Cambridge Institute of Public Health,
        Forvie Site, Robinson Way,
        Cambridge Biomedical Campus,
        Cambridge CB2 0SR,
        United Kingdom
    }\\
    MRC Biostatistics Unit \\
    University of Cambridge \\
    \texttt{kevin.kunzmann@mrc-bsu.cam.ac.uk} \\
    Michael J. Grayling \\
    Population Health Sciences Institute \\
    Newcastle University \\
    \texttt{michael.grayling@newcastle.ac.uk} \\
    Kim May Lee \\
    Pragmatic Clinical Trials Unit\\
    Queen Mary University of London\\
    \texttt{k.m.lee@qmul.ac.uk} \\
    David S.\ Robertson \\
    MRC Biostatisics Unit\\
    University of Cambridge\\
    \texttt{david.robertson@mrc-bsu.cam.ac.uk} \\
    Kaspar Rufibach \\
    Methods, Collaboration, and Outreach Group (MCO) \\
    Department of Biostatistics\\
    F. Hoffmann-La Roche, Basel\\
    \texttt{kaspar.rufibach@roche.com} \\
    James M. S. Wason \\
    Population Health Sciences Institute \\
    Newcastle University\\
    and \\
     MRC Biostatistics Unit \\
    University of Cambridge \\
    \texttt{james.wason@newcastle.ac.uk}
}

@@

The three estimators are depicted in Figure~\ref{fig:acp-ocp-pp-comparison-absolute} as functions 
of the observed interim outcome, where the interim time-point $m = 26$.
\begin{figure}
    \centering
    \begin{subfigure}[b]{\textwidth}
        \centering
        \includegraphics[width=.9\textwidth]{figures/acp_ocp_pp_example}
        \caption{%
            Estimates as function of the interim data.
        }
        \label{fig:acp-ocp-pp-comparison-absolute}
    \end{subfigure}
    \begin{subfigure}[b]{\textwidth}
        \centering
        \includegraphics[width=.9\textwidth]{figures/acp_ocp_pp_sampling_distributions}
        \caption{%
            Histograms of the sampling distributions.
        }
        \label{fig:sampling-distributions}
    \end{subfigure}
    \begin{subfigure}[b]{\textwidth}
        \centering
        \includegraphics[width=.9\textwidth]{figures/acp_ocp_pp_bias_mae_mse}
        \caption{%
            Bias, mean absolute error (MAE) and mean squared error (MSE).
        }
        \label{fig:acp-ocp-pp-as-estimators}
    \end{subfigure}
    \caption{%
        Properties of $\ACP,\OCP$, and $\PP$ as estimators of the unknown conditional power at $m=26$ an overall sample size $n=79$.
    }
\end{figure}
The most sensitive measure is $\OCP$ while $\ACP$ is the least sensitive
(see also spread of the sampling distributions in Figure~\ref{fig:sampling-distributions}). 
This is due to $\ACP$ assuming a fixed effect size and thus 
only being affected by direct changes in $z_m$.
Both $\OCP$ and $\PP$, however, are also indirectly affected by updating 
the belief about the effect size with the interim results.
The plug-in estimate $\OCP$ assumes that the observed effect is the 
true effect and does not quantify the uncertainty around this value in any way. 
$\PP$, on the other hand, invokes Bayes' theorem and then integrates 
$\CP(z_m,c,\theta)$ with respect to the obtained posterior distribution. 
The degree of adaptation of $\PP$ thus depends 
on the vagueness of the chosen prior. 
In fact, as the prior approaches a point mass at $\theta_1$, $\PP$ converges
to $\ACP$.
$\ACP$ can thus be seen as a mere special case of $\PP$ with a point prior
on $\theta_1>0$.

@@

If the recalculation rule defined in \eqref{eq:naive-recalc}--\eqref{eq:nmax-naive-recalc} is made mandatory in the
study protocol, 
it yields two functions $n(z_m)$ and $c(z_m)$ as the point-wise solution of the problem
which jointly define an `adaptive design' for any $z_m$.
Here, `mandatory' means that it is decided \textit{a~priori} to always (for all $z_m$)
recalculate the 
final sample size and critical value in the 
above specified way after a fixed number of outcomes $m$ has been observed.
The term `adaptive design' is slightly misleading in this context.
The design itself is pre-specified (and thus not adapted or changed) 
but only $n$ and $c$ are adaptive since they vary as functions of $z_m$.
We would like to emphasize that such a detailed specification of the sample size recalculation rule is not common in practice since most protocols are focused on guaranteeing control of the maximal type-I-error rate control and maintaining maximal flexibility during the interim analysis.
Still, if a sample-size reassessment was prescribed in the study protocol, there must have been some rationale as to the goal of that reassessment and we merely investigate the consequences for a design's unconditional properties when applying such a rationale in a mandatory way. 
The described approach should thus be seen as a hypothetical example highlighting the ineffectiveness that can arise from a na\"{\i}ve implementation of a recalculation based on a fixed threshold for 
conditional power.
The corresponding sample size and critical value functions for $m=26$ are depicted in Figure~\ref{fig:binding-recalculation}.
\begin{figure}
    \centering
    \includegraphics[width=\textwidth]{figures/naive_adaptive_and_optimal_designs}
    \caption{%
        Original single-stage design; na\"{\i}vely adapted design 
        with sample size $n'(z_m)$ and critical value $c'(z_m)$ functions defined by the conditional optimization problem \eqref{eq:naive-recalc}--\eqref{eq:nmax-naive-recalc} and 
        ${n=79}, {m=26}, {n_{max}=160}, {n_{min}=30}$ and $1-\beta_{cond.}=0.8$;
        the optimal design is the solution of \eqref{eq:optimal-design-start}--\eqref{eq:optimal-design-end}.
    }
    \label{fig:binding-recalculation}
\end{figure}

@@

We illustrate both approaches by example of the optimal two-stage design that minimizes the objective 
$\E_\varphi[\,n(Z_m)\,]$ subject to a maximal type-I error rate of $0.025$ and a minimal expected power of $0.8$, see~Figure~\ref{fig:binding-recalculation}.  
This design has a pre-planned interim analysis after $m=35$ individuals.
Now assume that
the prior $\varphi$ needs to be revised due to new trial-external 
information at the pre-planned interim analysis.
We consider members of the class of shifted truncated Normal distributions 
\begin{align}
    \psi_{\mu}(\theta) := \frac{\displaystyle \boldsymbol{1}_{[-0.5, 1]}(\theta)\,\phi\left(\frac{\theta - \mu}{0.2}\right)}{\displaystyle \Phi\left(\frac{1 - \mu}{0.2}\right) - \Phi\left(\frac{-0.5 - \mu}{0.2}\right)}
\end{align}
as revised priors, i.e. the original prior~$\varphi=\psi_{0.4}$.
Figure~\ref{fig:recalculation-old-m} shows the behavior of the recalculated sample size and critical value in this scenario for observed effects of $z_m/\sqrt{m}=0.25, 0.3, 0.35$ and varying mean of the revised prior if the originally planned interim time-point is maintained.
\begin{figure}
    \centering
    \begin{subfigure}[b]{\textwidth}
        \centering
        \includegraphics[width=.9\textwidth]{figures/recalculation-type-II-original-m}
        \caption{%
            Interim analysis at the originally planned interim sample size of $m=35$ for varying observed effect.
        }
        \label{fig:recalculation-old-m}
    \end{subfigure}
    \begin{subfigure}[b]{\textwidth}
        \centering
        \includegraphics[width=.9\textwidth]{figures/recalculation-type-II-earlier}
        \caption{%
            Earlier-than-planned interim analysis at an observed effect of $0.3$ and varying interim time-point.
        }
        \label{fig:recalculation-earlier}
    \end{subfigure}
    \begin{subfigure}[b]{\textwidth}
        \centering
        \includegraphics[width=.9\textwidth]{figures/recalculation-lambda-approach-original-m}
        \caption{%
            Interim analysis at the originally planned interim sample size of $m=35$ for varying observed effect and the $\lambda$ approach.
        }
        \label{fig:recalculation-lambda}
    \end{subfigure}
    \caption{%
        Comparison of consistent unplanned recalculation schemes.
    }
\end{figure}
Clearly, the original design is indeed invariant if the prior remains unchanged.
The sample size is increased if the new prior is more conservative
than $\varphi$ and \textit{vice~versa}.
